Modelling the travel time of rtansit vehicles in \rt through a GTFS-based road network using GPS vehicle locations

### 2019-08-09 anzjs2018/sections/04_implementation.tex
More edits

The most time consuming component was ETA writing,
which involved summarising the individual ETAs estimated for each particle.
While not discussed here this involved computing quantiles which uses a sorting algorithm
with complexity $O(N \log N)$,
which explains non-linear relationship with $N$.

@@

\begin{figure}[tb]
    \centering
    \includegraphics[width=0.8\textwidth]{figures/04_model_results_timing.pdf}
    \caption{
        The average real-time running time of each step within each iteration
        over a 15~minute off-peak window.
        The vehicle update and writing ETA steps involve sorting the particles,
        which is an operation of $\mathcal{O}(N\log N)$ complexity.
        Total time is for the entire iteration, which includes fetching the data from the API.
        % Total time increases as the number of particles $N$ increases.
        % The reason for the large contribution of writing ETAs is that
        % this step involves estimation of quantiles,
        % which requires sorting the particles, 
        % The vehicle update step also involves sorting whenever the sample
        % degenerations and resampling is required.
        % Total time also includes fetching the data from the API,
        % which is constant for any $N$, which explains the increase
        % in standardised total time for $N=500$.
    }
    \label{fig:timings}
\end{figure}

@@

To evaluate the performance of the model,
the simulation was run using a range of values of GPS error, $\omega^2$,
and system noise, $\sigma^2$.
Figure \ref{fig:dist_to_route} shows the distribution of the distance
between each observation and the route using a nearest point algorithm,
which suggested using values $\omega \in \{1,2,3,5\}$ for GPS error in meters.
For the system noise we used values of $\sigma^2\in \{1e^{-4},1e^{-3},1e^{-2},0.05\}$,
which correspond to an average vehicle speed variation of between 0.1 and 45~meters per second
over a period of 30~seconds.
For each combination of these parameter values, including varying $N$,
the following values were calculated:
effective sample size, $N_\text{eff}$;
degeneration rate, the percentage of iterations in which the vehicle was lost
(no plausible particles);
and the (relative) variance of segment travel time estimates.

\begin{figure}[tb]
    \centering
    \includegraphics[width=0.8\textwidth]{figures/04_model_results_dist.pdf}
    \caption{
        The distribution of minimum distance between observed vehicle location
        and the route path, truncated to 20~meters.
    }
    \label{fig:dist_to_route}
\end{figure}

@@

\begin{figure}[tb]
    \centering
    \includegraphics[width=\textwidth]{figures/04_model_results_neff.pdf}
    \caption{
        The effective sample size, $N_\text{eff}$,
        and degeneration rate for varying values of GPS error,
        system noise, and number of particles $N$.
        % increases with GPS error and decreases with system noise,
        % however both effects are reduced for larger $N$.
        % Conversely, degeneration rate decreases with GPS error and
        % system noise, although the effect of the latter is very small,
        % and again the relationship is less pronounced for larger $N$.
    }
    \label{fig:perf_stats}
\end{figure}

% \begin{figure}[tb]
%     \centering
%     \includegraphics[width=0.9\textwidth]{figures/04_model_results_degen.pdf}
%     \caption{
%         The degeneration rate is the percentage of iterations where there were no
%         plausible particles, so the vehicle was ``lost'' by the model.
%         This decreases as GPS error and number of particles increase,
%         but there is no visible effect of system noise.
%         Note, however, that some of these can be attributed to invalid data,
%         for example if the driver has entered the wrong trip direction.
%         The error bars show $\pm 2$SE.
%     }
%     \label{fig:degen_rate}
% \end{figure}

The last measure of performance is the relative variance of segment travel times.
We used relative variance as each segment has a very different travel time distribution,
so to compare the variability between simulations we needed to 
compute the average variance of estimated travel times over all simulations,
taking the average (over road segments) of the ratio between variance for
each individual simulation and variance over all simulations.
If $\tilde z_\ell^{\omega,\epsilon}$ is a vector of all travel times
along segment $\ell$ during the simulation with GPS error $\omega$ and system noise $\epsilon$,
then the relative variance $\bar v^{\omega,\epsilon}$ is calculated by
\begin{equation*}
\begin{split}
v_\ell^{\omega,\epsilon} &=
\frac{\mathrm{Var}(\tilde z_\ell^{\omega,\epsilon})}{\mathrm{Var}(\tilde z_\ell)} \\
\bar v^{\omega,\epsilon} &=
    \frac{1}{L}\sum_{\ell=1}^L v_\ell^{\omega,\epsilon}
\end{split}
\end{equation*}
Figure~\ref{fig:travel_times} shows that simulations with larger GPS error
resulted in a higher relative variance,
while simulations with more particles had lower relative variance.
There was no visible effect of system noise on the variability of travel times.

The results of the simulations
demonstrate a tradeoff between particle filter performance
($N_\text{eff}$ and degeneration rate) and parameter estimation.
However, until the arrival time prediction model has been completed,
it is difficult to make decisions about the optimal values to use:
we are unable to say whether a higher variation of travel time estimates
is going to have a significant effect on the final ETAs,
particularly when compared to the uncertainty of forecasting road state,
or whether reducing the degeneration rate, 
or increasing $N_\text{eff}$, would be more advantageous.

>>>>

The most time consuming component was ETA writing,
which involved summarising the individual ETAs estimated for each particle.
While not discussed here this involved computing quantiles which uses a sorting algorithm
with complexity $O(N \log N)$,
which explains the non-linear relationship with $N$.

@@

\begin{figure}[tb]
    \centering
    \includegraphics[width=0.8\textwidth]{figures/04_model_results_timing.pdf}
    \caption{
        The average real-time running time of each step within each iteration
        over a 15~minute off-peak window.
        The vehicle update and writing ETA steps involve sorting the particles,
        which is an operation of $\mathcal{O}(N\log N)$ complexity.
        Total time is for the entire iteration, which includes fetching the data from the API.
    }
    \label{fig:timings}
\end{figure}

@@

To evaluate the performance of the model,
the simulation was run using a range of values of GPS error, $\omega^2$,
and system noise, $\sigma^2$.
Figure \ref{fig:dist_to_route} shows the distribution of the distance
between each observation and the route using a nearest point algorithm,
which suggested using values $\omega \in \{1,2,3,5\}$ for GPS error in meters.
For the system noise we used values of $\sigma^2\in \{1e^{-4},1e^{-3},1e^{-2},0.05\}$,
which correspond to an average vehicle speed variation of between 0.1 and 45~meters per second
over a period of 30~seconds.
For each combination of these parameter values, including varying $N$,
the following values were calculated:
effective sample size, $N_\text{eff}$;
degeneration rate, the percentage of iterations in which the vehicle was lost
(no plausible particles remained);
and the (relative) variance of segment travel time estimates.

\begin{figure}[tb]
    \centering
    \includegraphics[width=0.7\textwidth]{figures/04_model_results_dist.pdf}
    \caption{
        The distribution of minimum distance between observed vehicle location
        and the route path, truncated to 20~meters.
    }
    \label{fig:dist_to_route}
\end{figure}

\begin{figure}[tb]
    \centering
    \includegraphics[width=\textwidth]{figures/04_model_results_neff.pdf}
    \caption{
        The effective sample size, $N_\text{eff}$,
        and degeneration rate for varying values of GPS error,
        system noise, and number of particles $N$.
    }
    \label{fig:perf_stats}
\end{figure}

@@

The last measure of performance is the relative variance of segment travel times.
We used relative variance as each segment has a very different travel time distribution,
so to compare the variability between simulations we needed to 
compute the average variance of estimated travel times over all simulations,
taking the average (over road segments) of the ratio between variance for
each individual simulation and variance over all simulations.
If $\tilde \bz_\ell^{\omega,\sigma}$ is a vector of all travel times
along segment $\ell$ during the simulation with GPS error $\omega$ and system noise $\sigma$,
then the relative variance $\bar v^{\omega,\sigma}$ is calculated by
\begin{equation*}
\begin{split}
v_\ell^{\omega,\sigma} &=
\frac{\mathrm{Var}(\tilde \bz_\ell^{\omega,\sigma})}{\mathrm{Var}(\tilde \bz_\ell)} \\
\bar v^{\omega,\sigma} &=
    \frac{1}{L}\sum_{\ell=1}^L v_\ell^{\omega,\sigma}
\end{split}
\end{equation*}
Figure~\ref{fig:travel_times} shows that simulations with larger GPS error
resulted in a higher relative variance,
while simulations with more particles had lower relative variance.
There was no clear pattern to the effect of system noise.

The results of the simulations
demonstrate a tradeoff between particle filter performance
($N_\text{eff}$ and degeneration rate) and parameter estimation.
However, until the arrival time prediction model has been completed,
it is difficult to make decisions about the optimal values to use:
we are unable to say whether a higher variation of travel time estimates
is going to have a significant effect on the final ETAs,
particularly when compared to the uncertainty of forecasting road state.

### 2019-08-09 anzjs2018/sections/05_discuss.tex
More edits

The network model presented here is simple, 
but provides the foundation for future work to
develop a formal model combining \rt travel time information with historical data, 
allowing short- and medium-term forecasts of road state,
for example incorporating peak time congestion.
% This in turn will provide the basis for an arrival time prediction model
% that combines \rt vehicle state with travel time forecasts
% in order to provide commuters with more reliable ETAs.
As for the arrival times, 
we plan to estimate these from a combination of historical 
and \rt data.
The model presented in this paper provides the starting point
for a transit modelling and predictive framework
that can be adapted to specific transit providers,
combined with historical data,
and run in \rt to provide commuters with
improved ETAs with which they can more reliably
plan their commutes.

The results of our simulations show that the framework is a 
faesible \rt alternative.
While the timings presented were for off-peak times,
we hope futher investigation of choosing $N$ and $N_\text{eff}$,
as well as work removing some redudancy from the transition algorithm,
will mean times during peak periods, when there are 2-3~times
as many vehicles, will come in under our 30~second target.

>>>>

The network model presented here is simple, 
but provides the foundation for future work to
develop a formal model combining \rt travel time information with historical data, 
allowing short- and medium-term forecasts of road state,
for example incorporating peak time congestion.
As for the arrival times, 
we plan to estimate these from a combination of historical 
and \rt data.
The model presented in this paper provides the starting point
for a transit modelling and predictive framework
that can be adapted to specific transit providers,
combined with historical data,
and run in \rt to provide commuters with
improved ETAs with which they can more reliably
plan their commutes.

The results of our simulations show that the framework is a 
faesible \rt alternative.
While the timings presented were for off-peak times,
futher investigation into the choice of $N$ and $N_\text{eff}$,
as well as work removing some redudancy from the transition algorithm,
should allow us to model all vehicles during peak periods---%
when there are 2-3~times as many vehicles---%
within our 30~second timeframe target.
